Imputation of Incomplete Multilevel Data with R

### 2024-01-23 Manuscript/manuscript.qmd
Update

Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modeled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster. However, this strategy is best suited for balanced data (Grund, 2017) and when the number of clusters is adequate sufficient for the random effects model (Austin,2018).

@@

### Assess Estimation Procedure Robustness to Missing Data
Before diving into imputation, make sure your estimation model can handle missing data. Sometimes, simpler methods like complete case analysis might be suffice, especially if your missingness is low (usually <5%). There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{molenberghs2007}.

<!-- What does it mean if an estimation model is not able to handle missing data? there are models that implicitaly are able to manage the missingness like the FML where in the likelihood it is used the information of the completed observations and the incomplete ones.  What do you mean with estimation model? the main model you use to analyse the data-->
<!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results. The text appears to suggest that in that case ML based methods are better. what i understood is ML are better in that case but please check it.-->

@@

<!-- most of this subsection is also mentioned at the start of #sec-workflow -->

@@

\begin{tabular}{ll}
     \toprule
Type & Definition \\
  \midrule
-2  & Cluster variable, in this case the one defined by j index  \\
 1  & Fixed variable, e.g., level-1  $x_{1_{ij}}$ or  level-2 $x_{2_{i.}}$ \\
 2  & Random variable, e.g., $x_{3_{ij}}$\\
 3  & *Fixed variable with cluster mean $x_{4_{ij}}$\\
 4  & *Random variable with cluster mean $x_{5_{ij}}$\\
-3. & Random variable only included on selection model (Heckman model)\\
-4. & Random variable only included on main model (Heckman model)\\
      \bottomrule
\end{tabular}

<!-- Can we reorder this table? It starts from -2, then -1 and 0 are missing, 1 to 4 and then again 2 negative numbers. What about the following?  -->

@@

In this example, we demonstrate a multilevel imputation of random intercept and random slope model with a continuous response. We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. We simulate data for 5 clusters so that the true values are known. We use the following variables from the dataset:

<!-- What do you mean with "We simulate data for 5 clusters"? If this dataset is included in the package why do we need to simulate the data? -->

@@

We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration. An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
Nonetheless, with respect to precision of the estimates, we notice that in general assuming MNAR leads to wider confidence intervals. In this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on the significance of a statistical test and therefore lead to contradictory conclusions.

<!-- Contradictory? Do you mean if the results from an imputation assuming MAR and one assuming MNAR disagree? -->

@@

The "Mice" and "mice"-based R packages offer a range of imputation methods tailored for hierarchical data, easily adaptable to the dataset's structure. Before proceeding with the analysis of the imputed dataset, it is essential to assess the convergence of the imputation method. This evaluation can reveal issues such as circular problems, the need for additional iterations, or challenges associated with the chosen imputation method.

>>>>

Under this hierarchical imputation model, observations within clusters are correlated and this correlation is modelled by a random effect so the hierarchical model can be estimated even when there are few observations per cluster. However, this strategy is best suited for balanced data (Grund, 2017) and when the number of clusters is adequate sufficient for the random effects model (Austin,2018).

@@

### Assess Estimation Procedure Robustness to Missing Data
Before diving into imputation, make sure your main model can handle missing data. Sometimes, simpler methods like complete case analysis might be suffice, especially if your missingness is low (usually <5%). There are scenarios in which specific Maximum Likelihood (ML) estimation methods outperform Multiple Imputation (MI) methods, for instance when the response variable is the sole incomplete variable, mixed models demonstrate robustness to missing data under the Missing at Random (MAR) assumption and with a correct variance-covariance specification.\cite{molenberghs2007}.

<!-- What does it mean if an estimation model is not able to handle missing data? there are models that implicitely are able to manage the missingness like the FML where computation of the likelihood uses the information from the completed observations and also from the incomplete ones.  What do you mean with estimation model? the main model used to analyse the data-->
<!-- AFAIK, when only the outcome is missing, MI and a ML based method lead to similar results. The text appears to suggest that in that case ML based methods are better. what i understood is ML are better cause they lead to efficient results, user do not have to make a lot of choises and it provides a determinist result. check this: https://missingdata.org/why-multiple-imputation-is-better-than-maximum-likelihood/#:~:text=Maximum%20likelihood%20is%20faster%20and,every%20time%20you%20run%20it. -->

@@

<!-- most of this subsection is also mentioned at the start of #sec-workflow agree i dont know how Hanne wants to integrate this as she wrote the start section after we completed this section. -->

@@

@@

In this example, we demonstrate a multilevel imputation of random intercept and random slope model with a continuous response. We utilize the obesity dataset included in the `micemd`\@ package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. We use the following variables from the dataset:

@@

We note that there is minimal disparity in the age effect, FamObs, or  Gender across the various imputation models under consideration. An analysis of the intercept reveals that, under the MNAR assumption, a higher average BMI is anticipated compared to the MAR assumption.
Nonetheless, with respect to precision of the estimates, we notice that in general assuming MNAR leads to wider confidence intervals. In this case it does not have any influence on the final result but there could be cases where variation in the assumed missing mechanism could lead also to differences on the significance of a statistical test and therefore lead to different conclusions.

<!-- Contradictory? Do you mean if the results from an imputation assuming MAR and one assuming MNAR disagree? yes indeed -->

@@

The "Mice" and "Mice"-based R packages offer a range of imputation methods tailored for hierarchical data, easily adaptable to the dataset's structure. Before proceeding with the analysis of the imputed dataset, it is essential to assess the convergence of the imputation method. This evaluation can reveal issues such as circular problems, the need for additional iterations, or challenges associated with the chosen imputation method.
